Statically typed Web programming with JavaScript

### 2012-12-03 papers/ecoop2013/repairing.tex
On the way

@@

This paper continues their work and presents and discusses the implementation of the following contributions:

\begin{itemize}

% JSDom, Forest
\item We expose native JavaScript APIs for DOM manipulation and event handling as statically typed APIs while reducing their syntactic noise ;

% State monad, FutureOps, OptionOps
\item We define a high-level abstraction to handle program's state changes and an abstraction turning asynchronous computations into first-class values. Then we show how to exploit staging to translate these abstractions into JavaScript with a minimal overhead ;

% NumericOps, Typeclasses
\item We use Scala's advanced type-level constructs to elegantly express type coercion constraints on operands in arithmetic operations and to achieve \emph{adhoc} polymorphism with no overhead at all in the generated code ;

% Forest server-side, cross-browser incompatibilities
\item Our code generation system is flexible enough to handle cross-browsers incompatibilities and, unlike other approaches, to allow the sharing of concepts between client-side and server-side even if these concepts are translated into code using external APIs.

\end{itemize}

The next section reviews the framework used to define compiled embedded DSLs in Scala, sections \ref{improving} to \ref{lms-benefits} present our contributions and section \ref{discussion} concludes.

@@

\section{Background and motivating example}

@@

We think this code is fragile for two reasons. (1) The name of the event is passed as a string, so it is easy to mispell it. (2) The callback passed as a second parameter takes a parameter \code{e} whose fields vary according to the listened event but developers have no way to check that the fields they’re using are indeed defined on the event they’re listening to.

@@

\begin{lstlisting}
var createVertex = function (text) {
  Ajax.post("/create", { content: text }, function (vertex) {
    addVertex(vertex);
  });
}
\end{lstlisting}

@@

\begin{lstlisting}[language=JavaScript,label=async-js]
var createVertex = function (text, callback) {
  Ajax.get("/create", { content: text }, function (data) {
    callback(data);
  });
}

@@

However increasing the number of callbacks makes the code flow harder to follow and adds distance to the programmer initial intent.

@@

Notice that the code is getting deeper toward the right, this problem is often named “callback-hell”. This code is also hard to reason about: when will the second \code{console.log} statement be executed? Before or after \code{baz} has been fetched?

@@

Not only the \code{vertexDom} function is very verbose, but it does not help to see the markup structure.

@@

This implementation may be more readable, however the code is wrong: if we call it with a content containing brackets (\eg{} \code{"<foo>"}) they won’t be escaped and will produce a \code{foo} tag nested in the \code{text} tag, which is not the intended behavior. So, although this way reads better it’s not less error prone.

By the way, markup generation is a task that often needs to be performed from both server and client sides. From the server-side it produces HTML pages that search engines can crawl and from the client-side it produces DOM fragments that can be used to update the user interface. We want to share HTML templates on both server and client sides to get consistency in the rendering and to avoid to duplicate the markup structure.

HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}} aim to make simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML structure and allowing the insertion of dynamic expressions in safe way. However, template engines require templates to be compiled, so they add an extra step in the build chain. Some template engines can be used on both client-side and server-side but they have severe limitations (\eg{} Closure Template has no expression language) and are not statically typed.

@@

- Few static checks

- Code hard to reason about

- Code too much verbose

A few existing template engines can be used on both server and client sides. They provide a convenient syntax to describe a HTML structure in which you can inject dynamic values. However, these template engines are limited because they’re not statically typed and they don’t provide a convenient expression language for the dynamic values. These two problems are in fact two aspects of a same problem: if we write the expression \code{xs.size == 0} in a template, how should it be translated in JavaScript and in Scala? Several solutions are valid, depending on the meaning of the expression: in JavaScript we could translate it as \code{xs.size === 0} or \code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a collection is named \code{length}). So we need to define a proper expression language and to define how it should be translated into JavaScript and Scala, but such a task requires a high effort and is hard to achieve in an extensible way (so users can still 
use their own data types in expressions instead of being restricted to a set of supported types). That’s probably for this reason that existing template engines usable on server and client sides have no expression language at all: the dynamic content can only be provided as a map of key-value pairs and no operation can be done on them.

... so we need better ways to handle DOM definition, user interaction events and asynchronous programming.

\section{DSL design and implementation}

This section presents both the design and the implementation of a compiled embedded DSL addressing the problems introduced in the preceding section.

@@

We propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods turning them into first-class citizen. Our DSL is monadic so we can solve the callback-hell problem thanks to the Scala \code{for} notation.

@@

The \code{map} and \code{flatMap} methods allow asynchronous values to be transformed and passed around, thus turning them into first-class citizen.

@@

\begin{lstlisting}
for {
  foo <- Ajax.get(fooUrl())
  bar <- Ajax.get(barUrl(foo))
  _ <- future(println("bar = " + bar))
  baz <- Ajax.get(bazUrl(bar))
} yield println(foo + bar + baz)
\end{lstlisting}

The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the first Ajax request is available, the next statement will be executed, and so on.

(FIXME Update the following or skip?) The above listing produces the following code (we renamed some identifiers for the sake of readability):

\begin{lstlisting}[language=JavaScript]
var x1 = new Promise();
AjaxGet("/foo", x1);
var x2 = new Promise();
AjaxGet("/bar", x2);
x1.foreach(function (foo) {
  x2.foreach(function (bar) {
    var x3 = foo + bar;
    console.log(x3);
  });
});
\end{lstlisting}

@@

@@

@@

Then we add a \code{+} method to any \code{Rep[_]} value, using an implicit class:

\begin{lstlisting}
implicit class NumericOps(lhs: Rep[A]) {
  def + (rhs: Rep[A]) = numeric_plus(lhs, rhs)
}
\end{lstlisting}

@@

\begin{lstlisting}
def main(i: Rep[Int], j: Rep[Int],
         x: Rep[Double], y: Rep[Double]) = {
  println(i + j)
  println(x + y)
  // i is promoted to Rep[Double], n has type Rep[Double]
  val n = i + y
  // j is promoted to Rep[Double], m has type Rep[Double]
  val m = x + j
  println(n + m)
}
\end{lstlisting}

\subsection{Discussion}

@@

\section{Comparison with other approaches}

@@

jQuery deferred, Q.js. Previous scala-js paper.

\section{Conclusion and perspectives}

We exposed native JavaScript APIs as statically typed APIs while decreasing their verbosity.
More code sharing.

>>>>

\authorrunning{Julien \noun{Richard-Foy} \etal{}}

@@

This paper continues their work and presents safe and expressive DSL constructs for RIAs development.

% This paper continues their work and presents and discusses the implementation of the following contributions:
% 
% \begin{itemize}
% 
% % JSDom, Forest
% \item We expose native JavaScript APIs for DOM manipulation and event handling as statically typed APIs while reducing their syntactic noise ;
% 
% % State monad, FutureOps, OptionOps
% \item We define a high-level abstraction to handle program's state changes and an abstraction turning asynchronous computations into first-class values. Then we show how to exploit staging to translate these abstractions into JavaScript with a minimal overhead ;
% 
% % NumericOps, Typeclasses
% \item We use Scala's advanced type-level constructs to elegantly express type coercion constraints on operands in arithmetic operations and to achieve \emph{adhoc} polymorphism with no overhead at all in the generated code ;
% 
% % Forest server-side, cross-browser incompatibilities
% \item Our code generation system is flexible enough to handle cross-browsers incompatibilities and, unlike other approaches, to allow the sharing of concepts between client-side and server-side even if these concepts are translated into code using external APIs.
% 
% \end{itemize}

The next section introduces a simple application example showing challenges of RIA development, section \ref{solution} presents our DSL tackling these challenges, section \ref{validation} validates our solution and compares it to related work and section \ref{discussion} concludes.

@@

\section{Background and motivating example}
\label{problem}

@@

We think this code is fragile for two reasons. (1) The name of the event is passed as a string, so it is easy to mispell it. (2) The callback passed as a second parameter takes a parameter \code{e} whose fields vary according to the listened event but developers have no way to check that the fields they’re using are indeed defined on the event they’re listening to. For instance, in our case we use the \code{wheelDeltaY} event field that is defined only on the \code{mousewheel} event.

@@

\begin{lstlisting}
var createVertex = function (text) {
  Ajax.post("/create", { content: text }, function (vertex) {
    addVertex(vertex);
  });
};
\end{lstlisting}

@@

\begin{lstlisting}[language=JavaScript,label=async-js]
var createVertex = function (text, callback) {
  Ajax.get("/create", { content: text }, function (data) {
    callback(data);
  });
};

@@

However increasing the number of callbacks makes the code flow harder to follow and adds distance to the programmer's initial intent.

@@

Notice that the code is getting deeper toward the right, this problem is often referred to as the “callback-hell”. This code is also hard to reason about: when will the second \code{console.log} statement be executed? Before or after \code{baz} has been fetched?

@@

Not only the \code{vertexDom} function is very verbose, but it does not reflect the markup nested structure, making it hard to read and reason about.

@@

This implementation may be more readable, however the above implementation is wrong: if we call it with a content containing brackets (\eg{} \code{"<foo>"}) they won’t be escaped and will produce a \code{foo} tag nested in the \code{text} tag, which is not the intended behavior. So, although this way reads slightly better it’s not less error prone.

By the way, unlike the previous points markup generation is a task that often needs to be performed from both server and client sides. From the server-side it produces HTML pages that search engines can crawl and from the client-side it produces DOM fragments that can be used to update the user interface. So we want to share HTML fragments definition between both server and client sides to get consistency in the rendering and to avoid duplication.

HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}} aim to make simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML structure and allowing the insertion of dynamic expressions in safe way. Some template engines can be used on both client-side and server-side, however none has a practical expression language (dynamic content can only be provided as a map of key-value pairs and no operation can be done on values) or is statically typed.

@@

In the previous section we showed the three main tasks performed by the JavaScript code in RIAs: events handling, asynchronous programming and DOM manipulation. We noticed the following issues:

\subsubsection{Lack of static checks}

Dynamic typing may give programmers more flexibility but can make it harder for little script to grow into mature and robust code bases and to perform refactorings.

\subsubsection{Code hard to reason about}

Callback-driven APIs are very common in JavaScript but their inverted control makes the code flow hard to follow.

\subsubsection{Verbose native APIs}

Native APIs are too much verbose, making the code longer than it should be.

\subsubsection{Limited sharing of code between client and server sides}

The lack of static checks is even harder to tackle in the case of code shared between client and server sides. If we write the expression \code{xs.size == 0} in a template, how should it be translated in JavaScript and in Scala? Several solutions are valid, depending on the meaning of the expression: in JavaScript we could translate it as \code{xs.size === 0} or \code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a collection is named \code{length}). We need to define a proper expression language and to define how it should be translated into JavaScript and Scala, but such a task requires a high effort and is hard to achieve in an extensible way (so users can still use their own data types in expressions instead of being restricted to a set of supported types).

\section{DSL design and implementation}
\label{solution}

This section presents both the design and the implementation of a compiled embedded DSL addressing the problems introduced in the previous section.

@@

We propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods turning them into first-class citizen. The DSL is monadic so we can solve the callback-hell problem thanks to the Scala \code{for} notation.

@@

Asynchronous values can be transformed using the \code{map} and \code{flatMap} methods, turning them into first-class citizen: functions can take as parameters and return asynchronous values.

@@

\begin{lstlisting}
for {
  foo <- Ajax.get(fooUrl())
  bar <- Ajax.get(barUrl(foo))
  _ <- future(println("bar = " + bar))
  baz <- Ajax.get(bazUrl(bar))
} println(foo + bar + baz)
\end{lstlisting}

The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the first Ajax request is available, the next statement will be executed, and so on. There is no nested callbacks and the order of execution is directly reflected by the order of statements.

The following listing shows the generated JavaScript code for an example where two asynchronous computations run in parallel (we renamed some identifiers for the sake of readability):

\begin{multicols}{2}
\begin{lstlisting}
val fooAsync = Ajax.get("/foo")
val barAsync = Ajax.get("/bar")
for {
  foo <- fooAsync
  bar <- barAsync
} println(foo + bar)
\end{lstlisting}
\vfill
\columnbreak
\begin{lstlisting}[language=JavaScript]
var x1 = new Promise();
AjaxGet("/foo", x1);
var x2 = new Promise();
AjaxGet("/bar", x2);
x1.onComplete(function (foo) {
 x2.onComplete(function (bar) {
  var x3 = foo + bar;
  console.log(x3);
 });
});
\end{lstlisting}
\end{multicols}

@@

(TODO More details about this optimization)

@@

(TODO More general problem description)

@@

Then we add a \code{+} method to any \code{Rep[_]} value, using an implicit class, so we can use an homogeneous syntax to write arithmetic operations whatever the numeric type of values this operation is applied on:

\begin{lstlisting}
implicit class NumericOps[A : Numeric](lhs: Rep[A]) {
  def + (rhs: Rep[A]) = numeric_plus(lhs, rhs)
}
\end{lstlisting}

@@

\begin{lstlisting}
def main(i: Rep[Int], j: Rep[Int],
         x: Rep[Double], y: Rep[Double]) = {
  println(i + j)
  // i is promoted to Rep[Double], m has type Rep[Double]
  val m = i + y
  // j is promoted to Rep[Double], n has type Rep[Double]
  val n = x + j
  println(m + n)
}
\end{lstlisting}

\subsection{Implementation discussion}

We implemented our language as a compiled embedded DSL in Scala. The embedding gave us the opportunity to reuse Scala’s infrastructure: we didn’t have to write a parser and, more importantly, we could reuse Scala’s type system. Furthermore, staging gave us the opportunity to generate efficient code.

Code generation using LMS is usually described as a two step process: a program first evaluates to an intermediate representation and then the final program’s code is generated from the intermediate representation. We think we can consider a third step, occurring before the program evaluation: the compilation.

Indeed, the code generation process consists in three steps giving DSL authors as many opportunities to implement their language features. First, features like type coercion and \emph{adhoc} polymorphism only rely on the type system, so they operate during the compilation. Then, during the evaluation, the DSL constructs build an intermediate representation of the program, for instance the State monad DSL builds an imperative sequence of assignments on a variable. Last, the code generator produces the final program using the domain specific information available in the intermediate representation to generate efficient code, for example the Option monad code generator produces code checking against nullity before dereferencing a value.

Each of these steps takes out some information on the program.

@@

(FIXME Simplify this subsubsection)

\subsubsection{Lack of a data type definition DSL}

Defining a language as an embedded DSL gives benefits but on the other side of the coin some features are more difficult to achieve. For instance there is no simple way to implement a DSL construct for defining data types.

(TODO Explain briefly why)

\section{Validation and comparison with other approaches}
\label{validation}

Our DSL aims to give developers a safer and more expressive way to write the client-side code. We focused on the main tasks of RIAs development: handling events, asynchronous programming and DOM definition.

\subsection{Rewrite of the application using the DSL?}

@@

Several JavaScript libraries turn asynchronous values into first-class citizen (\eg{} jQuery.Deferred\footnote{\href{http://api.jquery.com/category/deferred-object/}{http://api.jquery.com/category/deferred-object/}}, Q.js\footnote{\href{https://github.com/kriskowal/q}{https://github.com/kriskowal/q}}). They allow developers to register callbacks in order to be notified when the value is available.

Roy solves the callback-hell problem.

\section{Conclusion and perspectives}
\label{discussion}

We exposed native JavaScript APIs as statically typed APIs while decreasing their verbosity.

More code sharing.

\section{Acknowledgements}

We thank Tiark Rompf for his useful comments.
